---
title: "prepare"
author: "Yuchen Shao"
date: "2023-02-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
There are five frequently used `classes` of data: 1. numeric, 2. integer, 3. character, 4. logical, 5. complex.
```{r}
my_numeric <- 42
my_integer <- 2L #adding an L automatically denotes an integer
my_character <- "universe"
my_logical <- FALSE
my_complex <- 2+4i
```
You can use the `is()` and `as()` functions to clarify or specify a type of data.
Create `harry_potter_matrix` using the `matrix()` command. We need to tell R how to organize the `box_office` vector using the `nrow` and `byrow` commands.
```{r}
harry_potter_matrix <- matrix(box_office, nrow = 8, byrow = T)
harry_potter_matrix
```
Name the columns using `colnames()` with the vector region.
```{r}
colnames(harry_potter_matrix) <- region
```

Name the rows using `rownames()` with the vector titles.
```{r}
rownames(harry_potter_matrix) <- titles
```
And even add a new column to reflect this calculation. `cbind()` adds columns.
```{r}
all_harry_potter_matrix <- cbind(all_harry_potter_matrix_1, global)
all_harry_potter_matrix
```
Let's use lowercase names when we create the data frame. We just changed to lowercase here, but we could use any names we wish.  
```{r}
hbirds <- data.frame(sex = Sex, length_in = Length, weight_oz = Weight) #renaming will become more helpful in later labs
names(hbirds)
```
We can also select values in an entire column using the `$` sign. This can be useful for calculations. Here we calculate the mean of all of the birds weights. 
```{r}
w <- hbirds$weight_oz
mean(w)
```
We should save our hbirds data frame so we can use it again later! There are many ways to save data in R, here we write our data frame to a csv file. We use `row.names = FALSE` to avoid row numbers from printing out. 
```{r}
write.csv(hbirds, "hbirds_data.csv", row.names = FALSE)
```
In the previous part of the lab, you exported a `.csv` of hot springs data. Let's try to reload that `.csv`.  
```{r}
hot_springs <- readr::read_csv("hsprings_data.csv")
```
`head()` prints the first n rows of the data frame.
```{r}
#head(fish, n = 10)
```

`tail()` prinst the last n rows of the data frame.
```{r}
#tail(fish, n = 10)
```

`table()` is useful when you have a limited number of categorical variables. It produces fast counts of the number of observations in a variable. We will come back to this later... 
```{r}
table(fish$lakeid)
```
To add a range of columns use `start_col:end_col`.
```{r}
fish
```

```{r}
select(fish, fish_id:length)
```

Options to select columns based on a specific criteria include:  
1. ends_with() = Select columns that end with a character string  
2. contains() = Select columns that contain a character string  
3. matches() = Select columns that match a regular expression  
4. one_of() = Select columns names that are from a group of names 
To select all columns that are *not* a class of data, you need to add a `~`.
```{r}
select_if(fish, ~!is.numeric(.))
```
4. Rename any columns that have capitol letters or punctuation issues.  
```{r}
mammals <- select_all(mammals, tolower)
mammals<-select_all(mammals, ~str_replace(., " ", "_"))
names(mammals)
#colnames(mammals)[10]<-"afr"
#colnames(mammals)[11]<-"max_life"
#colnames(mammals)[3]<-"genus"
#names(mammals)
```

To keep things tidy, I am going to rename the variables in the mammals data.  
```{r}
mammals <- rename(mammals, genus=Genus, wean_mass="wean mass", max_life="max. life", litter_size="litter size", litters_per_year="litters/year")
```
You can also use `filter()` to extract data based on multiple conditions. Below we extract only the fish that have lakeid "AL" and length >350.
```{r}
filter(fish, lakeid == "AL" & length > 350)
```

Notice that the `|` operator generates a different result.
```{r}
filter(fish, lakeid == "AL" | length > 350)
```
## `mutate()`  
Mutate allows us to create a new column from existing columns in a data frame. We are doing a small introduction here and will add some additional functions later. Let's convert the length variable from cm to millimeters and create a new variable called length_mm.  
```{r}
fish %>% 
  mutate(length_mm = length*10) %>% 
  select(fish_id, length, length_mm)
```

## `mutate_all()`
This last function is super helpful when cleaning data. With "wild" data, there are often mixed entries (upper and lowercase), blank spaces, odd characters, etc. These all need to be dealt with before analysis.  

Here is an example that changes all entries to lowercase (if present).  
```{r}
mammals
```

```{r}
mammals %>%
  mutate_all(tolower)
```

Using the across function we can specify individual columns.
```{r}
mammals %>% 
  mutate(across(c("order", "family"), tolower))
```
```{r}
mammals %>% 
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))%>% 
  arrange(newborn)
```
Check out the way I am loading these data. If I know there are NAs, I can take care of them at the beginning. But, we should do this very cautiously. At times it is better to keep the original columns and data intact.  
```{r}
superhero_info <- readr::read_csv("data/heroes_information.csv", na = c("", "-99", "-"))
superhero_powers <- readr::read_csv("data/super_hero_powers.csv", na = c("", "-99", "-"))
```
```{r}
library("janitor")
```

The `clean_names` function takes care of everything in one line! Now that's a superpower!
```{r}
superhero_info <- janitor::clean_names(superhero_info)
names(superhero_info)
```
## `tabyl`
The `janitor` package has many awesome functions that we will explore. Here is its version of `table` which not only produces counts but also percentages. Very handy! Let's use it to explore the proportion of good guys and bad guys in the `superhero_info` data.  

```{r}
names(superhero_info)
tabyl(superhero_info, alignment)
```
```{r}
penguins %>% 
  tabyl(species, island) %>% 
  adorn_percentages() %>%
  adorn_pct_formatting(digits = 2)
```
```{r}
penguins %>%
  summarize(across(c(species, island, sex), n_distinct))
```
Here we summarize across all variables.
```{r}
penguins %>%
  summarise_all(mean, na.rm=T)
```
5. Based on the asfis_species_number, how many distinct fish species were caught as part of these data?
```{r}
fisheries_tidy%>%
  summarise(fish_species=n_distinct(asfis_species_number))
```
8. Which five countries caught the most cephalopods between 2008-2012?
```{r}
fisheries_tidy %>%
  group_by(country) %>%
  filter(isscaap_taxonomic_group=="Squids, cuttlefishes, octopuses") %>%
  filter(between(year, 2008, 2012)) %>%
  summarise(cephalopods_catch=sum(catch, na.rm=T)) %>%
  arrange(desc(cephalopods_catch))%>%
  head(n=5)
```
## Using `filter()` with multiple observations  
Filtering multiple values within the same variable requires the `%in%` [operator](https://www.tutorialspoint.com/r/r_operators.htm).    
```{r}
filter(fish, length %in% c(167, 175))
```

Alternatively, you can use `between` if you are looking for a range of specific values.  
```{r}
filter(fish, between(scalelength, 2.5, 2.55))
```

You can also extract observations "near" a certain value but you need to specify a tolerance.  
```{r}
filter(fish, near(radii_length_mm, 2, tol = 0.2))
```
Here we can be more efficient and summarize the number of NA's over all of the variables.
```{r}
msleep %>% 
  summarise_all(~(sum(is.na(.))))
```
Let's say we know that NA's are represented by "-999" in `amniota`. We can replace this value with NA across the data frame. You might be tempted to overwrite `amniota`, but in these cases I like to make a new object.
```{r}
amniota_tidy <- amniota %>% 
  na_if("-999")
```
`miss_var_summary` provides a nice, clean summary of NA's across the data frame.
```{r}
naniar::miss_var_summary(amniota_tidy)
```
Solution 1: specify a range of columns that you want to pivot.
```{r}
billboard2 <- 
  billboard %>% 
  pivot_longer(wk1:wk76, # a range of columns
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE #this will drop the NA's
               )
billboard2
```

Solution 2: OR, specify columns that you want to stay fixed.
```{r}
billboard3 <- 
  billboard %>% 
  pivot_longer(-c(artist, track, date.entered), #specific columns
               names_to = "week",
               values_to = "rank",
               values_drop_na = TRUE
               )
billboard3
```
Solution 3: identify columns by a prefix, remove the prefix and all NA's.
```{r}
billboard %>% 
   pivot_longer(
   cols = starts_with("wk"),
   names_to = "week",
   names_prefix = "wk",
   values_to = "rank",
   values_drop_na = TRUE)
```
`names_sep` helps pull these apart, but we still have "exp" and "rep" to deal with.  
```{r}
qpcr_untidy %>% 
  pivot_longer(
    exp1_rep1:exp3_rep3,
    names_to= c("experiment", "replicate"),
    names_sep="_",
    values_to="mRNA_expression"
  )
```
## Example 4: more than one value or observation in a row
```{r}
length_data %>% 
  transform(length = str_split(length, ";")) %>%
  unnest(length)
```
We need to start by separating the patient names from their sexes. `separate()` needs to know which column you want to split, the names of the new columns, and what to look for in terms of breaks in the data.
```{r}
heartrate2 %>% 
  separate(patient, into= c("patient", "sex"), sep = "_")
```
```{r}
heartrate3%>%
  unite(patient_sex,c(patient,sex),sep=" ")
```
We want to pivot the `edu_level` data such that there is only a single row per demographic. 
```{r}
edu_level %>% 
pivot_wider(names_from = (education_level), #new column names come from the education_level column
                values_from = c(mean_income, n)) #values come from two separate columns
```
```{r}
ggplot(data=homerange,mapping = aes(x=log10.mass,y=log10.hra))+geom_point()+geom_smooth(method = lm,se=F)
```
In big data sets with lots of overlapping values, over-plotting can be an issue. `geom_jitter()` is similar to `geom_point()` but it helps with over plotting by adding some random noise to the data and separating some of the individual points.
```{r}
ggplot(data=homerange, mapping=aes(x=log10.mass, y=log10.hra))+
  geom_jitter()
```
Unlike `geom_bar()`, `geom_col()` allows us to specify an x-axis and a y-axis.
```{r}
homerange %>% 
  filter(family=="salmonidae") %>%
  select(common.name, log10.mass) %>% 
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_col()
```
`stat="identity"` allows us to map a variable to the y-axis so that we aren't restricted to counts.
```{r}
homerange %>% 
  filter(family=="salmonidae") %>% 
  ggplot(aes(x=common.name, y=log10.mass))+
  geom_bar(stat="identity")
```
8. The column titled `INSTNM` is the institution name. We are only interested in the University of California colleges. Make a new data frame that is restricted to UC institutions. You can remove `Hastings College of Law` and `UC San Francisco` as we are only interested in undergraduate institutions.
```{r}
uc_campus<-colleges_clean%>% 
  filter_all(any_vars(str_detect(., pattern = "University of California")))
```

Remove `Hastings College of Law` and `UC San Francisco` and store the final data frame as a new object `univ_calif_final`.
```{r}
univ_calif_final<-uc_campus%>%filter(city!="San Francisco")
```
```{r}
#options(scipen = 999)#cancel scientific notation for the session
life_history %>% 
  group_by(order) %>% 
  summarize(mean_mass=mean(mass, na.rm=T))%>%
  ggplot(aes(x=reorder(order,mean_mass),y=mean_mass))+geom_col()+coord_flip()+scale_y_log10()
```

The `rel()` option changes the relative size of the title to keep things consistent. Adding `hjust` allows control of title position.
```{r}
elephants %>% 
  ggplot(aes(x=age, y=height)) + geom_point() + geom_smooth(method=lm, se=F) +
  labs(title = "Elephant Age vs. Height",
       x = "Age",
       y = "Height") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```
There are lots of options for aesthetics. An aesthetic can be assigned to either numeric or categorical data. `fill` is a common grouping option; notice that an appropriate key is displayed when you use one of these options.
```{r}
elephants%>%
  ggplot(aes(x=sex,fill=sex))+geom_bar()
```

`size` adjusts the size of points relative to a continuous variable.
```{r}
life_history%>%
  ggplot(aes(x=gestation,y=log10(mass),size=mass))+geom_point(na.rm = T)
```
Play with point size.
```{r}
p +geom_point(size=2)
```

Map shapes to another categorical variable.
```{r}
p+geom_point(aes(shape=thermoregulation, color=thermoregulation), size=1.75)
```
We can also have counts of each trophic guild within taxonomic group shown side-by-side by specifying `position="dodge"`.
Here is the same plot oriented vertically.
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Observations by Taxon in Homerange Data",
       x = "Taxonomic Group",
       fill= "Trophic Guild")
```
We can also scale all bars to a percentage (or proportion).
```{r}
homerange %>% 
  ggplot(aes(x = taxon, fill = trophic.guild))+
  geom_bar(position = position_fill())+ 
  scale_y_continuous(labels = scales::percent)+
  coord_flip()
```
In addition to `fill`, `group` is an aesthetic that accomplishes the same function but does not add color.

6. Add another layer to your answer from #5 using `geom_point` to get an idea of how many measurements were taken for each species.
```{r}
deserts %>% 
  filter(weight!="NA") %>% 
  ggplot(aes(x=species_id, y=weight)) +
  geom_point(alpha=0.2, color="lightblue", position = "jitter") +
  geom_boxplot(alpha=0.3)+
  coord_flip()+
  labs(title = "Distribution of weight for each species",
       x = "Species ID",
       y = "Weight")
```
```{r}
deserts2 %>% 
  filter(species_id=="DM" | species_id=="DS") %>% 
  group_by(year, species_id) %>% 
  summarise(n=n(), .groups='keep') %>% 
  ggplot(aes(x=year, y=n, group=species_id, color=species_id))+
  geom_line()+
  geom_point(shape=5)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Number of samples for species DM & DS",
       x = "Year",
       fill = "n")
```
What does the distribution of body mass look like in the `homerange` data?
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_histogram(alpha=0.4,color = "pink", fill = "lightskyblue3", bins=20)+
  labs(title = "Distribution of Body Mass")
```
Density plots are similar to histograms but they use a smoothing function to make the distribution more even and clean looking. They do not use bins.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_density(fill="deepskyblue4", alpha  =0.4, color = "black")+
  labs(title = "Distribution of Body Mass")
```
```{r}
homerange <- homerange %>% 
             mutate(mass_category = case_when(log10.mass <= 1.75 ~ "small",
                                              log10.mass > 1.75 & log10.mass <= 2.75 ~ "medium",
                                              log10.mass > 2.75 ~ "large"))
```
```{r}
p <- homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(na.rm=T, position="dodge")
p
```
```{r}
p+theme_classic()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
```{r}
p+theme_linedraw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
+`scale_colour_brewer()` is for points  
+`scale_fill_brewer()` is for fills  
Here I chose the `Paired` palette. Take a moment and experiment with other options.
```{r}
p+scale_fill_brewer(palette = "BuPu")+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
```{r}
p+scale_fill_manual(values=my_palette)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, color = locomotion)) +
  geom_point() +
  xlim(0, 4) +
  ylim(1, 6)
```
`facet_wrap()` makes a ribbon of panels, but you can control how you want them arranged.  
```{r}
ecosphere %>% 
  ggplot(aes(x=diet, y=log10_mass, fill=migratory_strategy))+ 
  geom_boxplot(alpha=0.4) + 
  facet_wrap(~migratory_strategy, ncol=6)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Mass by Diet and Migratory Strategy",
       x = NULL,
       y = "Log 10 Mass",
       fill = "Migratory Strategy")
```
`facet_grid()` allows control over the faceted variable; it can be arranged in rows or columns.  
rows~columns
```{r}
ecosphere %>% 
  ggplot(aes(x=diet, y=log10_mass, fill=migratory_strategy))+ 
  geom_boxplot(alpha=0.4)+ 
  facet_grid(migratory_strategy~.)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Mass by Diet and Migratory Strategy",
       x = NULL,
       y = "Log 10 Mass",
       fill = "Migratory Strategy")
```
`facet_grid()` will also allow the comparison of two categorical variables, just remember a~b where a is rows and b is columns.  
```{r}
ecosphere %>% 
  ggplot(aes(x=log10_mass))+
  geom_density()+
  facet_grid(diet~habitat, scales = "free_y")
```





